Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex
Further develop Technique and Results section under rating-based method.

Using this model, we may view the problem of estimating the similarity $Y$ of
items $A$ and $B$ that most likely produced ratings $R_A$ and $R_B$ as
that of finding the Maximum Likelihood Estimate of $Y | \vec{R}$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|\vec{R}=\vec{r}) 
\\&= \argmax_y\left[\frac{P(\vec{R}=\vec{r} |
Y=y)P(Y=y)}{P(\vec{R}=\vec{r})}\right]
\\&= \argmax_y\left[P(\vec{R}=\vec{r}|Y=y)P(Y=y)\right]
\\&=
\argmax_y\left[\frac{1}{2\pi\sqrt{\left|\mathbf{\Sigma_y}\right|}}\exp{\left(-\frac{1}{2}
(\vec{r} - \vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r})\right)}
\frac{1}{\sigma_{s}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu_s)^2}
{2\sigma_{s}^2}\right)}\right]
\\&= \argmin_y\left[(\vec{r} -
\vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r}) + \frac{(y -
\mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&=
\argmin_y\left[\vec{r^\prime}^\mathrm{T}\mathbf{\Sigma_y}^{-1}\vec{r^\prime}
 + \frac{(y - \mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&= \argmin_y\left[\frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right],
\end{align}
where we used the fact that
\begin{align}
\mathbf{\Sigma_y}^{-1} = \frac{1}{\sigma_r^2}
\left[ \begin{array}{cc}  1 & y \\ y & 1 \end{array} \right]^{-1} =
\frac{1}{\sigma_r^2}\frac{1}{1 - y^2}\left[ \begin{array}{cc}  1 & -y \\ -y & 1
\end{array} \right].
\end{align}
and that
\begin{align}
\left|\mathbf{\Sigma}\right| = \sigma_r^2(1 - y^2)
\end{align}
along with the substitution $\vec{r^\prime} = (r_A^\prime~~r_B^\prime) = \vec{r}
- \vec{\mu_r}$. Next, we find the minimum by taking the derivative with respect
to $y$ and setting it to zero:
\begin{align}
\frac{d}{dy}\left[ \frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right] &= 0 \\
-r_A^\prime r_B^\prime \hat{y}^2 + ({r_A^\prime}^2 + {r_B^\prime}^2)\hat{y} - r_A^\prime
r_B^\prime + \frac{\sigma_r^2}{\sigma_s^2}(1 - \hat{y}^2)^2(\hat{y} - \mu_s) - \sigma_r^2(1
- \hat{y}^2)\hat{y} &= 0.
\end{align}
This final expression implicitly yields the MLE of $y$ as a function of the
ratings $r_A$ and $r_B$. That is, by constraining the parameters $\mu_r$,
$\sigma_r$, $\mu_s$, and $\sigma_s$ to realistic values (e.g. $\sigma_r,
\sigma_s > 0$), this expression defines a surface that can be used to
efficiently estimate $y$ given a pair of ratings\footnote{Note that this
equation neglects the constraint that $-1 \leq y \leq 1$. As a consequence, some
combinations of ratings and parameters can yield MLEs for $y$ that exceed $1$ in
magnitude. This is easily fixed, however, by applying this additional
constraint.}.

@@

As already mentioned, the estimator $PearSim$ is unable to say much about the
similarity of two items when only provided ratings from a single common user. In
this case, $PearSim$ yields $\pm1$ (or the indeterminate value $0/0$ in the case
that both ratings are zero) depending on whether the ratings have the same sign
or not. The value of $\hat{y}$ provided by the above model rectifies this by
generating similarity estimates in the range $[-1, +1]$ given a single pair of
ratings. We therefore consider each user who rates a pair of items as implicitly
predicting the similarity of the items. A key discovery of the present work is
that the accuracy of users' predictions in a training set is correlated with the
accuracy in a test set. We exploit this finding by ascribing a quantity we call
{\em user predictivity} to each user defined as
\begin{align}
p_u = e^{-\epsilon_u/\lambda},
\end{align}
where $\lambda$ is a parameter to be learned and
\begin{align}
\epsilon_u = \frac{1}{\left|I_u\right|}
\sum_{A,B \in I_u}{\left|\hat{y}(r_{u,A},r_{u,B}) - GoldSim(A, B)\right|}
\end{align}
with $I_u$ the set of item pairs rated by user $u$. We are once again using
$GoldSim$ as a proxy for $TrueSim$ (See {\bf Experiment} section) such that
$\epsilon_u$ is the average absolute error of user $u$'s similarity predictions.

@@

Having chosen values for $\mu_s$, $\sigma_s$, $\mu_r$, and $\sigma_r$ we are now
able to estimate the similarity of item pairs. We do this for each common user
$u$ in the training set in order to determine their average absolute error
$\epsilon_u$. Next, we apply supervised learning techniques to optimize the
value of the parameter $\lambda$ that appears in the definition of the {\em user
predictivity} $p_u$. We do this by minimizing the error in the similarity
estimates made against the training set where each estimate is the weighted
average of the common user's implied prediction. That is, our new estimate for
the similarity of items $A$ and $B$ is
\begin{align}
UserPredSim(A,B) = \frac{\sum_{u \in U_{AB}}{p_u\cdot\hat{y}(r_{u,A}, r_{u,B})}}
{\sum_{u \in U_{AB}}{p_u}},
\end{align}
where, as before, $U_{AB}$ is the set of common users for items $A$ and $B$.

>>>>

Using this model, we may view the problem of estimating the similarity $Y$ of
items $A$ and $B$ that most likely produced ratings $R_A$ and $R_B$ as
that of finding the Maximum Likelihood Estimate of $Y | \vec{R}$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|\vec{R}=\vec{r}) 
\\&= \argmax_y\left[\frac{P(\vec{R}=\vec{r} |
Y=y)P(Y=y)}{P(\vec{R}=\vec{r})}\right]
\\&= \argmax_y\left[P(\vec{R}=\vec{r}|Y=y)P(Y=y)\right]
\\&=
\argmax_y\left[\frac{1}{2\pi\sqrt{\left|\mathbf{\Sigma_y}\right|}}\exp{\left(-\frac{1}{2}
(\vec{r} - \vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r})\right)}
\frac{1}{\sigma_{s}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu_s)^2}
{2\sigma_{s}^2}\right)}\right]
\\&= \argmin_y\left[(\vec{r} -
\vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r}) + \frac{(y -
\mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&=
\argmin_y\left[\vec{r^\prime}^\mathrm{T}\mathbf{\Sigma_y}^{-1}\vec{r^\prime}
 + \frac{(y - \mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&= \argmin_y\left[\frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right],
\end{align}
where we used the fact that
\begin{align}
\mathbf{\Sigma_y}^{-1} = \frac{1}{\sigma_r^2}
\left[ \begin{array}{cc}  1 & y \\ y & 1 \end{array} \right]^{-1} =
\frac{1}{\sigma_r^2}\frac{1}{1 - y^2}\left[ \begin{array}{cc}  1 & -y \\ -y & 1
\end{array} \right].
\end{align}
and that
\begin{align}
\left|\mathbf{\Sigma}\right| = \sigma_r^2(1 - y^2)
\end{align}
along with the substitution $\vec{r^\prime} = (r_A^\prime~~r_B^\prime) = \vec{r}
- \vec{\mu_r}$. Next, we find the minimum by taking the derivative with respect
to $y$ and setting it to zero:
\begin{align}
\frac{d}{dy}\left[ \frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right] &= 0 \\\label{eqn:estimator}
-r_A^\prime r_B^\prime \hat{y}^2 + ({r_A^\prime}^2 + {r_B^\prime}^2)\hat{y} - r_A^\prime
r_B^\prime + \frac{\sigma_r^2}{\sigma_s^2}(1 - \hat{y}^2)^2(\hat{y} - \mu_s) - \sigma_r^2(1
- \hat{y}^2)\hat{y} &= 0.
\end{align}
This final expression implicitly yields the MLE of $y$ as a function of the
ratings $r_A$ and $r_B$. That is, by constraining the parameters $\mu_r$,
$\sigma_r$, $\mu_s$, and $\sigma_s$ to realistic values (e.g. $\sigma_r,
\sigma_s > 0$), this expression defines a surface that can be used to
efficiently estimate $y$ given a pair of ratings\footnote{Note that this
equation neglects the constraint that $-1 \leq y \leq 1$. As a consequence, some
combinations of ratings and parameters can yield MLEs for $y$ that exceed $1$ in
magnitude. This is easily fixed, however, by applying this additional
constraint.}.

@@

As already mentioned, the estimator $PearSim$ is unable to say much about the
similarity of two items when only provided ratings from a single common user. In
this case, $PearSim$ yields $\pm1$ (or the indeterminate value $0/0$ in the case
that both ratings are zero) depending on whether the ratings have the same sign
or not. The value of $\hat{y}$ provided by the above model rectifies this by
generating similarity estimates in the range $[-1, +1]$ given a single pair of
ratings. We therefore consider each user who rates a pair of items as implicitly
predicting the similarity of the items. A key discovery of the present work is
that the accuracy of users' predictions in a training set is correlated with the
accuracy in a test set. We exploit this finding by ascribing a quantity we call
{\em user predictivity} to each user defined as
\begin{align}\label{eqn:predictivity}
p_u = e^{-\epsilon_u/\lambda},
\end{align}
where $\lambda$ is a parameter to be learned and
\begin{align}\label{eqn:userError}
\epsilon_u = \frac{1}{\left|I_u\right|}
\sum_{A,B \in I_u}{\left|\hat{y}(r_{u,A},r_{u,B}) - GoldSim(A, B)\right|}
\end{align}
with $I_u$ the set of item pairs rated by user $u$. We are once again using
$GoldSim$ as a proxy for $TrueSim$ (See {\bf Experiment} section) such that
$\epsilon_u$ is the average absolute error of user $u$'s similarity predictions.

@@

Having chosen values for $\mu_s$, $\sigma_s$, $\mu_r$, and $\sigma_r$ we are now
able to estimate the similarity of item pairs. To do this, we must solve Eq.
(\ref{eqn:estimator}) for $\hat{y}$ given a pair of ratings. This can be easily
done in practice using a root finding algorithm\footnote{To solve Eq.
(\ref{eqn:estimator}) for $\hat{y}$ given ratings $r_A$ and $r_B$, we use the
{\tt brentq} root finding function of the widely available Python module {\tt
scipy.optimize}.}. The implied predictions of individual users can then be
aggregated to form a new estimate of the similarity of items $A$ and $B$ as
\begin{align}\label{eqn:userPredSim}
UserPredSim(A,B) = \frac{\sum_{u \in U_{AB}}{p_u\cdot\hat{y}(r_{u,A}, r_{u,B})}}
{\sum_{u \in U_{AB}}{p_u}},
\end{align}
where, as before, $U_{AB}$ is the set of common users for items $A$ and $B$ and
$p_u$ is the {\em user predictivity} defined in Eq. (\ref{eqn:predictivity}). We
leverage the training set to find $p_u$ for each user $u$. This is done by
recording the implied prediction $\hat{y}(r_{u,A},r_{u,B})$ for each common user
$u$ for each item pair $(A, B)$ in the training set and then computing each
user's average absolute error $\epsilon_u$ as per Eq. (\ref{eqn:userError}).
Finally, we apply supervised learning techniques to optimize the value of the
parameter $\lambda$ that appears in the definition of $p_u$. This is achieved by
minimizing the error in Eq. {\ref{eqn:userPredSim} made against the training
set.
